Constraints on Assembly Bias from Galaxy Clustering

### 2016-06-17 ms.tex
Minor edits to Interpretation.

The results of the previous section suggest that halo mass is insufficient to predict 
the galaxy content of a halo to the precision mandated by an analysis of SDSS DR7 
clustering measurements (particularly the $\magr < -19$, $\magr < -20$, and $\magr < -20.5$ samples). 
This should not be surprising, as the process of galaxy formation and evolution is complex and 
there is little reason to suspect that many environmental factors determine galaxy properties. 
Our results further suggest that halo concentration (or some other halo property correlated 
with halo concentration) impacts the galaxy content of a halo, at least for galaxies of some 
luminosities. In particular, the assembly bias inferred from the $\magr < -20$ and $\magr < -20.5$ 
samples suggests that galaxies in roughly Milky Way-sized halos tend to be more luminous 
if they reside in halos with higher-than-average concentrations.

This can be interpreted in at least two ways that are not mutually exclusive within the context of the standard 
model of structure formation. Our analysis cannot support either of these scenarios unambiguously, 
but we considered these reasonable starting points for further consideration. 
One possibility is simply that the growth and structure of the host 
halo itself may drive this dependence. For example, halos of higher concentration at fixed 
mass have deeper potential wells (specifically, higher escape velocities from the halo center). 
Therefore, gravity more strongly binds the stellar and gaseous contents of such halos possibly 
driving more rapid star formation or making the galaxies less susceptible to feedback mechanisms 
that suppress star formation. Higher concentration halos have also formed earlier, on average. This 
suggests that any physics that acts upon the baryonic content of the halo its high-density 
environment may have been operative for a longer time, resulting in a larger, more luminous galaxy. 
Halo assembly bias persists to large separations ($r \gg \mathrm{Mpc}$) and this picture can 
accommodate galaxy assembly bias on such large scales.

Another possibility is that halo-halo interactions in dense environments may drive, in part, 
this preference for more luminous galaxies in higher-concentration halos. Recent work 
on halo evolution suggests that a more physical definition of the size of a halo should be 
$\sim 2-3$ times the traditional virial radius 
\citep{wetzel_etal14,adhikari_etal14,wetzel_nagai15,sunayama_etal16,more_etal16}. 
This ``splashback radius" incorporates ``splashback" halos that have passed within the dense environment 
of the central halo, but are traditionally classified as distinct halos despite being strongly 
influenced by the environment of the central halo. \citet{wang_etal07}, \citet{wetzel_etal14},  
\citet{sunayama_etal16}, all find that these ``splashback" halos are a significant cause 
of halo assembly bias on scales $r \sim 1-3\, h^{-1}\mathrm{Mpc}$, where assembly bias is 
most pronounced and strongly scale dependent. Consequently, decorated HOD models 
exhibit this same feature \citet{hearin_etal16}. It is not a coincidence that the greatest 
improvement of the decorated HOD in the description of galaxy clustering occurs on these same 
length scales (see Fig.~\ref{fig:Mr19samples}-\ref{fig:Mr21samples} and the associated discussion).

This ``splashback" hypothesis has further merit in that it has the same sense as suggested 
by the data. Splashback halos have experienced mass loss and been spatially truncated 
as a consequence of their interactions with a larger, nearby halo. Therefore, splashback halos 
have masses that may be smaller than the masses of other halos that host similar galaxies \citet{dalal_etal08}. 
Moreover, high concentrations due to tidal truncation are a hallmark of splashback 
halos \citep{wang_etal07,dalal_etal08,sunayama_etal16}. This results in a picture in which there is a population 
of halos, correlated on scales $r \sim 1-3\, h^{-1}\mathrm{Mpc}$, that have concentrations 
that are high for their mass and galaxies that are overluminous for their mass. This is precisely the 
sense of assembly bias suggested by our analysis. Galaxy assembly bias induced by these 
splashback effects cannot explain larger-scale assembly bias.

@@

Until such time as we have stringent constraints on the level of assembly bias in the 
universe, accounting for the possibility of assembly bias is not optional in interpreting 
galaxy clustering data. Precise statements about galaxy evolution derived from 
galaxy clustering or attempts to exploit clustering to perform cosmological analyses 
cannot be made robustly without accounting for assembly bias. Together with 
the work of \citet{hearin_etal16} and the {\tt Halotools} software, we have provided 
an explicit example of one way forward in such analyses. It is our hope that similar analyses 
will be performed on forthcoming galaxy survey data and that these analyses will lead to a
richer understanding of galaxy evolution and the relationship between galaxies and the halos 
and environments in which they are found.

>>>>

The results of the previous section suggest that halo mass is insufficient to predict 
the galaxy content of a halo to the precision mandated by an analysis of SDSS DR7 
clustering measurements (particularly the $\magr < -19$, $\magr < -20$, and $\magr < -20.5$ samples). 
This should not be surprising, as the process of galaxy formation and evolution is complex and 
there is little reason to suspect that all potentially-important factors are determined by a single halo property. 
Our results further suggest that halo concentration (or some other halo property correlated 
with halo concentration) impacts the galaxy content of a halo, at least for galaxies of some 
luminosities. In particular, the assembly bias inferred from the $\magr < -20$ and $\magr < -20.5$ 
samples suggests that galaxies in roughly Milky Way-sized halos tend to be more luminous 
if they reside in halos with higher-than-average concentrations.

This can be interpreted in at least two ways that are not mutually exclusive within the context of the standard 
model of structure formation. Our analysis cannot support either of these scenarios unambiguously, 
but we considered these reasonable starting points for further consideration. 
One possibility is simply that the growth and structure of the host 
halo itself may drive this dependence. For example, halos of higher concentration at fixed 
mass have deeper potential wells (specifically, higher escape velocities from the halo center). 
Therefore, gravity more strongly binds the stellar and gaseous contents of such halos, possibly 
driving more rapid star formation or making the galaxies less susceptible to feedback mechanisms 
that suppress star formation. Higher concentration halos have also formed earlier, on average. This 
suggests that any physics that acts upon the baryonic content of the halo and its high-density 
environment may have been operative for a longer time, resulting in a larger, more luminous galaxy. 
Concentration-dependent halo assembly bias persists to very large separations 
\citep[$r > 10\, h^{-1}\mathrm{Mpc}$, e.g.,][]{wechsler06} 
and this picture can accommodate galaxy assembly bias on such large scales.

Another possibility is that halo-halo interactions in dense environments may drive, in part, 
this preference for more luminous galaxies in higher-concentration halos. Recent work 
on halo evolution suggests that a more physical definition of the size of a halo should be 
$\sim 2-3$ times the traditional virial radius 
\citep{wetzel_etal14,adhikari_etal14,wetzel_nagai15,sunayama_etal16,more_etal16}, 
a scale on the order of $\sim \mathrm{Mpc}$. 
This ``splashback radius" incorporates ``splashback" halos that have passed within the dense environment 
of a larger, central halo, but are traditionally classified as distinct halos despite being strongly 
influenced by the environment of the central halo. \citet{wang_etal07}, \citet{wetzel_etal14},  
\citet{sunayama_etal16}, all find that these ``splashback" halos are a significant cause 
of halo assembly bias on scales $r \sim 1-3\, h^{-1}\mathrm{Mpc}$, where assembly bias is 
most pronounced and strongly scale dependent. Consequently, decorated HOD models 
exhibit this same feature \citet{hearin_etal16}. It is not a coincidence that the greatest 
improvement of the decorated HOD in the description of galaxy clustering occurs on these same 
length scales (see Fig.~\ref{fig:Mr19samples}-\ref{fig:Mr21samples} and the associated discussion).

This ``splashback" hypothesis has further merit in that it has the same sense suggested 
by the data. Splashback halos have experienced mass loss and have been spatially truncated. 
Therefore, splashback halos have virial masses that may be smaller than the masses of other 
halos that host similar galaxies \citet{dalal_etal08}. Meanwhile, high concentrations due 
to tidal truncation are a hallmark of splashback halos 
\citep{wang_etal07,dalal_etal08,sunayama_etal16}. 
The implication is a population 
of halos, correlated on scales $r \sim 1-3\, h^{-1}\mathrm{Mpc}$, that have concentrations 
that are high for their mass and galaxies that are overluminous for their mass. This is precisely the 
sense of assembly bias suggested by our analysis. 
Galaxy assembly bias induced by these 
splashback effects cannot explain larger-scale assembly bias.

Our formulation of the Decorated HOD is not designed to address 
the finer-grained causes for observed correlations. 
Rather, what we have provided in this work is a robust boundary condition 
that more detailed specific models, such as semi-analytical models 
and hydrodynamical simulations, should satisfy. 
However, we reiterate an important corollary to our findings. 
Although SDSS projected clustering measurements favor strong levels of assembly bias, 
the constraints on even our simple $\abias$ parameters are relatively weak. 
This implies that {\em it may not be possible to constrain a finer-grained physical 
picture using projected clustering at a level that is interesting.} 
In order to constrain models that are more complex than simple empirical models 
such as the Decorated HOD, it is necessary to leverage additional constraining power of 
observations beyond two-point clustering.

@@

Until such time as we have stringent constraints on the level of assembly bias in the 
universe, accounting for the possibility of assembly bias is not optional in interpreting 
galaxy clustering data. Precise statements about galaxy evolution derived from 
galaxy clustering or attempts to exploit clustering to perform cosmological analyses 
cannot be made robustly without accounting for assembly bias \citep[see][]{zentner_etal14}. 
Together with the work of \citet{hearin_etal16} and the {\tt Halotools} software, we have provided 
an explicit example of one way forward in such analyses. It is our hope that similar analyses 
will be performed on forthcoming galaxy survey data and that these analyses will lead to a
richer understanding of galaxy evolution and the relationship between galaxies and the halos 
and environments in which they are found.
